---
title: Robert Williams, a.k.a. TimeLord, Has Emerged and Could Very Well Win DPOY Next Season 
author: Trey Fritz
date: '2022-04-25'
slug: robert-williams-a-k-a-timelord-has-emerged-and-the-rest-of-the-nba-should-be-worried
categories: []
tags: []
---

```{r warning=FALSE, message=FALSE, echo=FALSE}
library(tidyverse)
library(scales)
library(cowplot)
library(ggbeeswarm)
library(ggrepel)
library(ggbump)
library(waffle)
```

Robert Williams was drafted in 2018 with the 27th pick in the 1st round by the Boston Celtics. It was a bit of a rough start for Williams, missing his flight to Boston after the draft lead him to missing his first practice. After this slip up a few of his fellow Celtics decided to give him the nickname TimeLord, and it has stuck ever since. 

After missing that flight things started to get much better for Rob, while his numbers through his first couple years didn't blow anyone away he would make his presence known in the small amount of minutes he was getting. Fast forward a few years and Williams is one of the best centers in the league. He has put on pounds of muscle while never losing his insane jumping ability. With a wingspan of 7'5 1/2 and a 6'9 frame Williams is not easy to get past, he clogs up the lane more that a baby shoving toys into the toilet. 

This year TimeLord has taken a huge leap defensively and offensvively, but we are going to focus on his defense. His number of blocks, defensive rebounds and steals have all jumped up. Most importantly his blocks have gone up from 1.3 to 2.2, thats adding another block every game. 

Looking at a very important stat for defensive players now, defensive win shares, Williams has gone from .6 to 3.9. A defensive win share is the number of wins a player contributes due to their defense, so in short Rob is winning the Celtics games with his defense. His overall defensive win share ranks 5th in the league, the leader of this stat is his superstar teammate Jaysun Tatum. Besides Tatum, Williams is right behind Rudy Gobert, and I mean RIGHT behind. Gobert is considered one of if not the best defensive player in the league and has won defensive player of the year three times. 
```{r warning=FALSE, message=FALSE, echo=FALSE}
defplay <- read_csv("Untitled spreadsheet - Sheet1.csv") %>% mutate(DWS=as.numeric(DWS)) %>% rename(BLKpct = "BLK%") %>% mutate(BLKpct=as.numeric(BLKpct)) %>% mutate(DBPM=as.numeric(DBPM)) 

pos <- defplay %>% filter(Pos == "C" | Pos == "PF" | Pos == "SF")

set.seed(1234)


rw <- pos %>% filter(Player == "Robert Williams")
jt <- pos %>% filter(Player == "Jayson Tatum")

ggplot() + geom_beeswarm(data = pos, aes(x=Pos, y=DWS), color="grey") +
  geom_beeswarm(data = rw, aes(x=Pos, y=DWS), color="dark green") +
  geom_beeswarm(data = jt, aes(x=Pos, y=DWS), color="dark green") +
  geom_text_repel(data = rw, aes(x=Pos, y=DWS, label=Player), nudge_x = .25) +
  geom_text_repel(data = jt, aes(x=Pos, y=DWS, label=Player)) +
  theme_minimal() + 
  labs(title = "TimeLord is Near The Top in One of The Most Important Statistics", subtitle= "Williams may not be leading in DWS but his presence goes beyond the numbers.", x="", y="Defensive Win Shares",  caption="Source: Nba Reference | By Trey Fritz") +
  theme_minimal() +
  theme(
    plot.title.position = "plot",
    plot.title = element_text(size = 15, face = "bold"),
    plot.subtitle = element_text(size = 10) )
```
So Rob is in good company, his stats are starting to speak for themselves and it is only a matter of time before people start to catch on and give him some recognition, and some consideration for DPOY. 

While having a good defensive win share is great Williams is also starting to get near the top in a few other key stats, defensive box plus minus and block percentage. Defensive box plus minus is an estimate of the defensive points a player contributes every 100 possessions above a league-average player, and the block percentage is an estimate on how many two point shots are blocked while a player in on the floor. 
```{r warning=FALSE, message=FALSE, echo=FALSE}
dbpm <- pos %>% 
  filter(as.numeric(G) > 55) %>%  
  filter(as.numeric(Age) < 30) %>% 
arrange(desc(DBPM)) %>%
  top_n(5, wt=DBPM)

blks <- pos %>% 
  filter(as.numeric(G) > 55) %>%  
  filter(as.numeric(Age) < 30) %>% 
  arrange(desc(BLKpct)) %>% 
  top_n(5, wt=BLKpct)


 bar1 <- ggplot() + geom_bar(data = dbpm, aes(x=reorder(Player, DBPM), weight=DBPM), fill="grey") + 
  geom_bar(data = rw, aes(x=reorder(Player, DBPM), weight=DBPM), fill="dark green") + 
  labs(title = "Williams is Always Showing up at The Top", subtitle= "At such a young age Rob is pushing some established veterans for one of the highest DBPM.", x="", y="Defensive Box Plus Minus",  caption="") + coord_flip() +
  theme_minimal() +
  theme(
    plot.title.position = "plot",
    plot.title = element_text(size = 18, face = "bold"),
    plot.subtitle = element_text(size = 10) )
 

bar2 <- ggplot() + geom_bar(data = blks, aes(x=reorder(Player, BLKpct), weight=BLKpct),fill="grey") + 
  geom_bar(data = rw, aes(x=reorder(Player, BLKpct), weight=BLKpct), fill="dark green") + 
  labs(title = "", subtitle= "", x="", y="Block Percentage",  caption="Source: Sports Reference | By Trey Fritz") + coord_flip() +
  theme_minimal() 

plot_grid(bar1, bar2, ncol = 2)
```
While Rob isn't at the top in these categories he still has yet to hit his prime. Once he does and gets another year or so under his belt I wouldn't be suprised if Williams is leading in both of these categories. 

Some of you may still be doubting that Williams is about to emerge as one of the best defensive players in the league. So for comparison, and to show how good TimeLord is playing, I put the number of blocks and steals Williams had up against what three time DPOY Rudy Gobert has. 

Gobert barely beats out Williams in blocks, having just three more, but Williams has 11 more steals. Gobert is five years older than Rob, five years, and Rob is putting up similar numbers now. His growth will continue and he will become an absolute menace to opposing offenses. Next year Robert Williams is going to make a run for DPOY, and don't be surprised if he comes out on top. 
```{r warning=FALSE, message=FALSE, echo=FALSE}
rw <- c("Blocks"=134, "Steals"=56)
rg <- c("Blocks"=137, "Steals"=45)

iron(waffle(rw, rows = 8, title = "Robert Williams", xlab = "1 square = 1 Block/Steal", colors = c("dark green", "light grey")),
waffle(rg, rows = 8, title = "Rudy Gobert", xlab = "1 square = 1 Block/Steal", colors = c("navy", "grey"))) 

```

The code for this article can be found here. 






